---
title: "Mock PMX Data: Three Worked Examples"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Mock PMX Data: Three Worked Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%"
)
library(pmxmock)
```

## Purpose

This vignette applies `pmxmock` to three datasets from `nlmixr2data`. The goal
is to create **mock data for model-workflow exploration** without fitting or
calling a PK, PD, or NLME model.

The output is not anonymous data and has no formal privacy guarantee. It is
not appropriate for parameter estimation, scientific inference, model
selection, or clinical decisions.

The examples illustrate three event structures:

1. repeated oral dosing with one PK endpoint (`theo_md`);
2. separate PK and PD endpoints (`warfarin`); and
3. infusion start/stop events with a delayed biomarker trajectory (`wbcSim`).

## Shared workflow

Every example follows the same four steps:

1. declare column roles explicitly with `pmx_roles()`;
2. generate subjects with `mock_pmx()`;
3. validate event structure with `validate_pmx()`; and
4. inspect structural summaries and trajectories with `compare_pmx()`.

The helper below loads a named dataset and labels every plot with the concrete
source object and generator.

```{r helpers}
load_nlmixr2_data <- function(name) {
  environment <- new.env(parent = emptyenv())
  utils::data(
    list = name,
    package = "nlmixr2data",
    envir = environment
  )
  get(name, envir = environment, inherits = FALSE)
}

label_comparison <- function(comparison, source_name) {
  caption <- paste0(
    "Source: nlmixr2data::", source_name,
    " | Mock: pmxmock::mock_pmx()"
  )
  comparison$plots <- lapply(comparison$plots, function(plot) {
    plot + ggplot2::labs(caption = caption)
  })
  comparison
}
```

## Theophylline: repeated dosing

`theo_md` uses uppercase column names, has one concentration endpoint, and
contains seven dose records per subject. Dose and observation records can be
tied at the same nominal time.

```{r theo-generate}
theo_md <- load_nlmixr2_data("theo_md")

theo_roles <- pmx_roles(
  id = "ID",
  time = "TIME",
  dv = "DV",
  amt = "AMT",
  evid = "EVID",
  cmt = "CMT",
  covariates = "WT"
)

theo_mock <- mock_pmx(
  theo_md,
  theo_roles,
  seed = 101
)

theo_validation <- validate_pmx(theo_mock, theo_roles)
theo_validation$valid
```

The source and mock datasets have the same subject count because
`n_subjects = NULL` by default. Each generated subject receives a new numeric
ID.

```{r theo-structure}
theo_comparison <- compare_pmx(theo_md, theo_mock, theo_roles)
theo_comparison <- label_comparison(theo_comparison, "theo_md")
knitr::kable(theo_comparison$summary)

c(
  source_subjects = length(unique(theo_md$ID)),
  mock_subjects = length(unique(theo_mock$ID)),
  doses_per_mock_subject = unique(
    vapply(split(theo_mock$EVID, theo_mock$ID), function(x) sum(x != 0), integer(1))
  )
)
```

The overlay shows the overall concentration-time envelope. The faceted view
shows individual trajectories separately for source and mock data.

```{r theo-overlay, echo=FALSE}
theo_comparison$plots$overlay
```

```{r theo-individual, echo=FALSE}
theo_comparison$plots$faceted
```

The dose records themselves remain sampled templates; only IDs, subject-level
covariates, and genuine observation DVs are synthesized.

```{r theo-first-subject}
first_theo_id <- unique(theo_mock$ID)[1]
utils::head(theo_mock[theo_mock$ID == first_theo_id, ], 14)
```

## Warfarin: separate PK and PD endpoints

`warfarin` uses lower-case names. `dvid` is a factor with `cp` for plasma
concentration and `pca` for pharmacodynamic response. The endpoints must never
be blended together.

```{r warfarin-generate}
warfarin <- load_nlmixr2_data("warfarin")

warfarin_roles <- pmx_roles(
  id = "id",
  time = "time",
  dv = "dv",
  amt = "amt",
  evid = "evid",
  dvid = "dvid",
  covariates = c("wt", "age", "sex")
)

warfarin_mock <- mock_pmx(
  warfarin,
  warfarin_roles,
  seed = 202
)

warfarin_validation <- validate_pmx(warfarin_mock, warfarin_roles)
warfarin_validation$valid
levels(warfarin_mock$dvid)
levels(warfarin_mock$sex)
```

```{r warfarin-structure}
warfarin_comparison <- compare_pmx(
  warfarin,
  warfarin_mock,
  warfarin_roles
)
warfarin_comparison <- label_comparison(warfarin_comparison, "warfarin")
knitr::kable(warfarin_comparison$summary)
```

The endpoint facets make the very different PK and PD scales explicit. Each
endpoint has its own transformation, interpolation, blend, and noise process,
while both use the same subject-level donor weights.

```{r warfarin-overlay, echo=FALSE}
warfarin_comparison$plots$overlay
```

```{r warfarin-individual, echo=FALSE}
warfarin_comparison$plots$faceted
```

The endpoint-specific transformations are recorded with the result.

```{r warfarin-settings}
attr(warfarin_mock, "pmx_settings")$endpoint_transforms
```

## WBC: infusion and delayed response

`wbcSim` uses positive `AMT` and `RATE` records to start an infusion and
negative records to stop it. Those paired records are copied together from an
event template rather than synthesized numerically.

```{r wbc-generate}
wbc_sim <- load_nlmixr2_data("wbcSim")

wbc_roles <- pmx_roles(
  id = "ID",
  time = "TIME",
  dv = "DV",
  amt = "AMT",
  evid = "EVID",
  cmt = "CMT",
  rate = "RATE",
  covariates = c("V2I", "V1I", "CLI")
)

wbc_mock <- mock_pmx(
  wbc_sim,
  wbc_roles,
  seed = 303
)

wbc_validation <- validate_pmx(wbc_mock, wbc_roles)
wbc_validation$valid
```

Some WBC dosing schedules are unique. In that case, `mock_pmx()` emits a
warning and uses the anchor as the sole measurement donor, with randomized
trajectory noise providing the perturbation. This avoids silently mixing
incompatible schedules.

```{r wbc-structure}
wbc_comparison <- compare_pmx(wbc_sim, wbc_mock, wbc_roles)
wbc_comparison <- label_comparison(wbc_comparison, "wbcSim")
knitr::kable(wbc_comparison$summary)

wbc_events <- wbc_mock[wbc_mock$EVID != 0, c(
  "ID", "TIME", "AMT", "RATE", "CMT", "EVID"
)]
utils::head(wbc_events, 8)
```

The trajectories retain delayed declines, nadirs, and recoveries by blending
observed WBC paths. The known Friberg model is not called.

```{r wbc-individual, echo=FALSE}
wbc_comparison$plots$faceted
```

```{r wbc-settings}
settings <- attr(wbc_mock, "pmx_settings")
settings$alignment
settings$compatible_event_groups
settings$warnings
```

## What is preserved—and what is not

The prototype preserves:

- source column names, order, practical classes, and factor levels;
- complete subject event templates and within-time row ordering;
- coherent `EVID`, `AMT`, `RATE`, `CMT`, and `DVID` semantics;
- missing-observation conventions from the sampled template; and
- subject-level constancy of declared baseline covariates.

It does not attempt to preserve:

- PK, PD, or NLME parameters;
- covariate-response relationships;
- the source distribution exactly;
- scientific suitability for model selection; or
- formal anonymity or privacy guarantees.

Use `validate_pmx()` and `compare_pmx()` to inspect structure, then review the
mock data under the privacy and governance rules that apply to the source.
