---
title: "Exploring Epsilon in pmxSynthData"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Exploring Epsilon in pmxSynthData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(pmxSynthData)
```

## Scope

This vignette is about the **differentially private alternative** engines, not
the default AVATAR method. Read "Privacy in pmxSynthData" first for when the DP
engines are the right choice (data that crosses a trust boundary). If your
synthetic data stays inside a trusted environment, use `synthesize_pmx()` and
ignore epsilon entirely.

## What epsilon buys

Epsilon is the one-person influence limit: smaller means stronger privacy and
more noise, larger means weaker privacy and less noise. There is no universal
default; an approved value must come from governance and threat modeling, not
from whichever number makes a plot look best.

For the structural-correction engine (`fit_calibrated_pmx()`), the usable
accuracy is captured by one quantity:

$$f \;=\; \frac{d}{\varepsilon N}$$

where `d` is the number of released quantities, `N` the number of subjects, and
`f` the fraction of the prior's width that survives as noise. The `pmx_preflight()`
helper reports it before any budget is spent:

```{r preflight}
priors <- pmx_priors(pk = pmx_prior(c(1 / 4, 4), source = "example"))
pmx_preflight(priors, epsilon = 0.5, n_subjects = 60)
```

The decision rule is not "is the error small" but "does the release beat the
prior": when `f` is near or above 1 the release conveys nothing the prior did
not, and generating from public inputs alone is strictly better.

## The measured frontier

Utility degrades sharply in small cohorts, and this is a property of
differential privacy, not of the implementation. Measured total fold-error on
clearance, for the structural engine with a correction-factor prior:

| N | epsilon 0.25 | epsilon 1 |
|---:|---:|---:|
| 6 | 2.3-fold | 1.7-fold |
| 20 | 1.6-fold | 1.3-fold |
| 60 | 1.4-fold | 1.1-fold |
| 300 | 1.08-fold | 1.06-fold |

From about 60 subjects upward the error is limited by estimator bias rather than
by the privacy mechanism, so a smaller epsilon is the better buy: it strengthens
the guarantee at no real cost in accuracy. The full derivation and measurements
are in `design/FEASIBILITY.md` and `design/PRIVACY_BACKGROUND.md`.

## Where to read more

- `design/PRIVACY_BACKGROUND.md` — the `d`, `f`, epsilon arithmetic in full.
- `design/FEASIBILITY.md` — the complete measured frontier by cohort size.
- `design/METHOD_DISCUSSION.md` — why AVATAR is the default and DP the
  boundary-crossing alternative.
