---
title: "Privacy in pmxSynthData"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Privacy in pmxSynthData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(pmxSynthData)
```

## The two methods, and when privacy matters

`pmxSynthData` offers two families of method. The **default** is AVATAR-style
blending (`synthesize_pmx()`), documented in the "Using pmxSynthData" and "How
pmxSynthData Works" vignettes. It resamples and blends whole source subject
trajectories, works at any cohort size, and makes **no formal privacy
guarantee**. It is intended for mock data used **inside a trusted computing
environment**, in the same spirit as Novartis's `synadam`.

This vignette describes the **alternative**: the differentially private engines,
for the case that needs a formal guarantee.

## The decision rule

The choice is a single question:

> **Does the generated data cross a trust boundary?**

- **Stays inside** the safe environment — you are the only consumer, governance
  and access controls apply → use **AVATAR** (`synthesize_pmx()`). Its lack of a
  formal guarantee costs nothing, because there is no adversary to guarantee
  against.
- **Crosses out** — shared with a partner or vendor, published, or moved to a
  less-controlled system → use a **differentially private** engine. A formal
  guarantee is the only thing that survives a determined adversary.

Differential privacy is expensive precisely because it defends against someone
who wants to break it. Buy it when the output will be handed to strangers; skip
it when nothing can reach the output.

## The differentially private engines

Two are provided, both making a subject-level `(epsilon, delta)` guarantee:
neighboring datasets differing by one complete subject produce nearly
indistinguishable output, so no one person's participation can be inferred from
the release.

- **`fit_calibrated_pmx()`** (structural correction). Asserts curve shape from a
  public structural model and privately calibrates only the exposure magnitude
  by a small correction factor. Viable at small cohorts. This is the
  recommended DP path for early-phase studies.
- **`fit_private_pmx()`** (dense grid). Releases a larger set of noised
  population summaries. Retained for large pooled corpora where its cost is
  affordable.

Both fail closed when the validated OpenDP backend is unavailable and never
substitute ordinary random noise:

```{r backend, eval=requireNamespace("opendp", quietly = TRUE)}
dp_backend_status()
```

## What the guarantee does and does not mean

- **Epsilon** is the one-person influence limit: smaller is stronger. It is not
  a re-identification probability.
- **Delta** is a small additive slack in the probability bound. It is not the
  fraction of unprotected patients.
- Differential privacy bounds the information attributable to one person's
  participation. It does **not** establish legal anonymity, authorize release,
  or secure the environment.

## Where to read more

- `design/PRIVACY_BACKGROUND.md` — how the arithmetic works: `d`, `f`, epsilon,
  and the error law, with worked examples.
- `design/FEASIBILITY.md` — measured utility of the DP engines by cohort size,
  and why small cohorts are hard for any formal method.
- `design/METHOD_DISCUSSION.md` — the full AVATAR-versus-DP tradeoff and the
  `synadam` comparison.
- The "Exploring Epsilon" vignette — how epsilon trades off against utility.
