---
title: "Exploring Epsilon in pmxSynthData"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    fig_width: 12
    fig_height: 8
vignette: >
  %\VignetteIndexEntry{Exploring Epsilon in pmxSynthData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(pmxSynthData)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_opendp <- isTRUE(dp_backend_status()$available)
can_explore <- has_nlmixr2data && has_ggplot2 && has_opendp
epsilon_values <- c(Small = 5, Medium = 50, Large = 500)
comparison_colours <- c(Source = "#1B6CA8", Synthetic = "#D95F02")
```

# What this vignette does

This vignette fits the same coarse private population generator at three
illustrative epsilon values and displays each value in a separate panel. For
every dataset, source observations appear above synthetic observations and use
the same colors as the practical demo.

The labels **Small**, **Medium**, and **Large** are relative only to this
prototype and its conservative high-dimensional vector sensitivity:

| Label | Epsilon | Interpretation in this vignette |
|---|---:|---|
| Small | 5 | Strongest privacy and usually the most distortion among these three fits |
| Medium | 50 | Less noise, but still visibly affected in small studies |
| Large | 500 | Weakest privacy and generally closest to the public-fixture fit |

These values are not recommendations. In particular, epsilon 50 and 500 would
usually represent weak privacy. An approved epsilon must come from governance,
threat modeling, and the full release context—not from choosing the most
attractive plot.

The practical demo's `epsilon = 3` or `5` is **not a utility comparison at that
epsilon**: those examples use the deliberately noiseless
`backend = "public"` because their source datasets are already public. That
backend records the illustrative allocation but does not apply DP noise and
makes no privacy claim. This vignette uses `backend = "opendp"`, so its epsilon
5 panels can look much worse even though the numeric argument is the same.

All source datasets here are already public. Fitting three mechanisms to the
same confidential dataset would create three releases and compose their
privacy losses; under the package's basic composition the combined epsilon in
this illustration would be 555. Do not reproduce this sweep on confidential
data without explicit authorization and accounting.

The privacy-noise draws are intentionally not seedable. A vignette rebuild may
therefore look different. Generation from each fitted model uses a fixed
ordinary seed, and the displayed cohort size is held at the public source count
to isolate changes in regimen, sampling, and trajectory summaries. The table
beside each plot still reports the noisy fitted subject-count release.

```{r backend-availability}
data.frame(
  dependency = c("nlmixr2data", "ggplot2", "OpenDP production adapter"),
  available = c(has_nlmixr2data, has_ggplot2, has_opendp)
)
```

If a dependency is unavailable, the explanatory text and code remain in the
vignette but formal-noise panels are skipped. The package never substitutes
ordinary R noise for OpenDP.

# Evaluation helpers

The configurations deliberately omit dose times, dose counts, intervals,
amounts, rates, infusion duration, endpoint grids, and occasion schedules.
Those quantities are inferred inside each privacy-accounted fit.

```{r epsilon-helpers, eval=can_explore}
epsilon_budget <- function() {
  pmx_budget_allocation(.10, .15, .15, .10, .50, 0)
}

load_public_dataset <- function(name) {
  environment <- new.env(parent = emptyenv())
  utils::data(list = name, package = "nlmixr2data", envir = environment)
  get(name, envir = environment, inherits = FALSE)
}

epsilon_case <- function(name) {
  source <- load_public_dataset(name)
  if (name == "theo_md") {
    roles <- pmx_roles(
      id = "ID", time = "TIME", dv = "DV", amt = "AMT",
      evid = "EVID", cmt = "CMT", covariates = "WT"
    )
    endpoints <- list(cp = pmx_endpoint(
      alignment = "dose_relative", transform = "log", shape = "occasion",
      cmt = 2
    ))
    bounds <- pmx_bounds(
      c(0, 170), list(cp = c(0, 30)), amt = c(0, 500),
      covariates = list(WT = c(40, 130))
    )
    design <- pmx_public_design(
      pmx_schema(source), dose_evid = 101, dose_cmt = 1
    )
    limits <- pmx_contribution_limits(40, 8, 8, 30, 11)
  } else if (name == "warfarin") {
    roles <- pmx_roles(
      id = "id", time = "time", dv = "dv", amt = "amt",
      evid = "evid", dvid = "dvid", covariates = c("wt", "age", "sex")
    )
    endpoints <- list(
      cp = pmx_endpoint("cp", "dose_relative", "log", "occasion"),
      pca = pmx_endpoint("pca", "study_time", "identity", "global")
    )
    bounds <- pmx_bounds(
      c(0, 144), list(cp = c(0, 25), pca = c(0, 120)),
      amt = c(0, 200),
      covariates = list(wt = c(40, 150), age = c(18, 100))
    )
    design <- pmx_public_design(pmx_schema(source), dose_evid = 1)
    limits <- pmx_contribution_limits(
      30, 2, 2, c(cp = 20, pca = 12), 12
    )
  } else if (name == "wbcSim") {
    roles <- pmx_roles(
      id = "ID", time = "TIME", dv = "DV", amt = "AMT",
      evid = "EVID", cmt = "CMT", rate = "RATE",
      covariates = c("V2I", "V1I", "CLI")
    )
    endpoints <- list(wbc = pmx_endpoint(
      alignment = "study_time", transform = "log", shape = "global", cmt = 3
    ))
    bounds <- pmx_bounds(
      c(0, 720), list(wbc = c(0, 30)), amt = c(-200, 200),
      rate = c(-200, 200),
      covariates = list(
        V2I = c(100, 1500), V1I = c(100, 1200), CLI = c(100, 800)
      )
    )
    design <- pmx_public_design(
      pmx_schema(source), dose_evid = 10101, dose_cmt = 1,
      endpoint_cmt = list(wbc = 3)
    )
    limits <- pmx_contribution_limits(20, 2, 2, 12, 9)
  } else stop("Unknown epsilon case.")
  list(
    name = name, source = source, roles = roles, endpoints = endpoints,
    bounds = bounds, design = design, limits = limits
  )
}

fit_epsilon_case <- function(case, epsilon, generation_seed) {
  model <- tryCatch(
    suppressWarnings(fit_private_pmx(
      case$source, case$roles, case$endpoints,
      epsilon = epsilon, delta = 0, bounds = case$bounds,
      public_design = case$design,
      contribution_limits = case$limits,
      budget_allocation = epsilon_budget(), backend = "opendp"
    )),
    error = function(error) error
  )
  if (inherits(model, "error")) {
    return(list(model = NULL, synthetic = NULL,
                error = conditionMessage(model)))
  }
  cohort_size <- length(unique(case$source[[case$roles$id]]))
  synthetic <- tryCatch(
    generate_pmx(
      model, n_subjects = cohort_size, seed = generation_seed
    ),
    error = function(error) error
  )
  if (inherits(synthetic, "error")) {
    return(list(model = model, synthetic = NULL,
                error = conditionMessage(synthetic)))
  }
  list(model = model, synthetic = synthetic, error = NULL)
}

observed_panel_data <- function(data, case, dataset, epsilon_label) {
  roles <- case$roles
  observed <- as.character(data[[roles$evid]]) %in% c("0", "0.0")
  if (!is.null(roles$mdv)) {
    observed <- observed & as.character(data[[roles$mdv]]) %in% c("0", "0.0")
  }
  observed <- observed & !is.na(data[[roles$dv]])
  rows <- which(observed)
  endpoint <- if (is.null(roles$dvid)) {
    rep(names(case$endpoints)[1L], length(rows))
  } else {
    declared <- vapply(
      case$endpoints, function(x) as.character(x$dvid), character(1)
    )
    names(declared) <- names(case$endpoints)
    names(declared)[match(as.character(data[[roles$dvid]][rows]), declared)]
  }
  out <- data.frame(
    dataset = factor(dataset, levels = c("Source", "Synthetic")),
    epsilon_label = factor(
      epsilon_label, levels = paste0(names(epsilon_values), " (", epsilon_values, ")")
    ),
    endpoint = factor(endpoint, levels = names(case$endpoints)),
    subject = as.character(data[[roles$id]][rows]),
    time = as.numeric(data[[roles$time]][rows]),
    dv = as.numeric(data[[roles$dv]][rows]), stringsAsFactors = FALSE
  )
  out[
    out$time >= case$bounds$time[1L] & out$time <= case$bounds$time[2L],
    , drop = FALSE
  ]
}

epsilon_status <- function(case, results) {
  do.call(rbind, lapply(seq_along(results), function(index) {
    result <- results[[index]]
    data.frame(
      setting = names(epsilon_values)[index],
      epsilon = unname(epsilon_values[index]),
      noisy_fitted_subject_count = if (is.null(result$model)) NA_real_ else
        result$model$population$private_subject_count,
      inferred_doses = if (is.null(result$model)) NA_integer_ else
        result$model$population$event$n_doses,
      generated_observations = if (is.null(result$synthetic)) NA_integer_ else
        sum(as.character(result$synthetic[[case$roles$evid]]) %in% c("0", "0.0")),
      result = if (is.null(result$error)) "generated" else result$error,
      stringsAsFactors = FALSE
    )
  }))
}

epsilon_panel <- function(case, results) {
  labels <- paste0(names(epsilon_values), " (", epsilon_values, ")")
  pieces <- list()
  failures <- list()
  for (index in seq_along(epsilon_values)) {
    label <- labels[index]
    pieces[[length(pieces) + 1L]] <- observed_panel_data(
      case$source, case, "Source", label
    )
    result <- results[[index]]
    if (!is.null(result$synthetic)) {
      pieces[[length(pieces) + 1L]] <- observed_panel_data(
        result$synthetic, case, "Synthetic", label
      )
    } else {
      failures[[length(failures) + 1L]] <- data.frame(
        dataset = factor("Synthetic", levels = c("Source", "Synthetic")),
        epsilon_label = factor(label, levels = labels),
        endpoint = factor(names(case$endpoints), levels = names(case$endpoints)),
        time = mean(case$bounds$time),
        dv = vapply(names(case$endpoints), function(endpoint) {
          mean(case$bounds$endpoints[[endpoint]])
        }, numeric(1)),
        label = paste("No generated table:\n", result$error),
        stringsAsFactors = FALSE
      )
    }
  }
  plotted <- do.call(rbind, pieces)
  line_group <- interaction(
    plotted$dataset, plotted$subject, plotted$endpoint,
    plotted$epsilon_label, drop = TRUE
  )
  line_data <- plotted[
    ave(seq_len(nrow(plotted)), line_group, FUN = length) > 1L,
    , drop = FALSE
  ]
  plot <- ggplot2::ggplot(
    plotted,
    ggplot2::aes(
      x = time, y = dv, colour = dataset,
      group = interaction(dataset, subject, endpoint, epsilon_label)
    )
  ) +
    ggplot2::geom_line(data = line_data, alpha = .28) +
    ggplot2::geom_point(alpha = .48, size = .55) +
    ggplot2::facet_grid(
      rows = ggplot2::vars(dataset, endpoint),
      cols = ggplot2::vars(epsilon_label), scales = "free_y", drop = FALSE
    ) +
    ggplot2::scale_colour_manual(values = comparison_colours, drop = FALSE) +
    ggplot2::labs(
      x = "Study time", y = "DV", colour = "Dataset",
      title = paste(case$name, "across three epsilon values"),
      subtitle = "Each column is a separate OpenDP fit; source is above synthetic"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "bottom")
  if (length(failures)) {
    failure_data <- do.call(rbind, failures)
    plot <- plot + ggplot2::geom_text(
      data = failure_data,
      ggplot2::aes(x = time, y = dv, label = label),
      inherit.aes = FALSE, colour = "#922B21", size = 3
    )
  }
  plot
}
```

# Theophylline

The Theophylline panels examine inferred repeated dosing, unusual
occasion-specific sampling, and a dose-relative PK-like trajectory. Dose rows
remain excluded from the plotted observations.

```{r theo-epsilon-fit, eval=can_explore}
theo_case <- epsilon_case("theo_md")
theo_results <- lapply(seq_along(epsilon_values), function(index) {
  fit_epsilon_case(theo_case, epsilon_values[index], 3100 + index)
})
knitr::kable(epsilon_status(theo_case, theo_results), digits = 2)
```

```{r theo-epsilon-panel, eval=can_explore, fig.height=7.5}
epsilon_panel(theo_case, theo_results)
```

# Warfarin

Warfarin has two clocks: dose-relative concentration (`cp`) and global
study-time pharmacodynamic response (`pca`). Each epsilon column contains both
endpoint rows.

```{r warfarin-epsilon-fit, eval=can_explore}
warfarin_case <- epsilon_case("warfarin")
warfarin_results <- lapply(seq_along(epsilon_values), function(index) {
  fit_epsilon_case(warfarin_case, epsilon_values[index], 3200 + index)
})
knitr::kable(epsilon_status(warfarin_case, warfarin_results), digits = 2)
```

```{r warfarin-epsilon-panel, eval=can_explore, fig.height=11}
epsilon_panel(warfarin_case, warfarin_results)
```

# WBC

The WBC panels examine a generalized infusion and a delayed study-time decline,
nadir, and recovery over the declared 0--720 hour comparison window.

```{r wbc-epsilon-fit, eval=can_explore}
wbc_case <- epsilon_case("wbcSim")
wbc_results <- lapply(seq_along(epsilon_values), function(index) {
  fit_epsilon_case(wbc_case, epsilon_values[index], 3300 + index)
})
knitr::kable(epsilon_status(wbc_case, wbc_results), digits = 2)
```

```{r wbc-epsilon-panel, eval=can_explore, fig.height=7.5}
epsilon_panel(wbc_case, wbc_results)
```

# How to interpret the panels

The expected tendency is better utility as epsilon increases, but a single
Laplace draw need not improve monotonically in every coordinate. Small epsilon
can collapse a sampling probability, saturate a bounded summary, change the
generalized number of doses, or make generation impossible; those outcomes are
shown rather than silently replaced. Large epsilon reduces noise but weakens
the privacy guarantee.

For a real evaluation, use `scripts/evaluate_simulations.R` across many public
generation seeds and, when authorized on public/simulated sources, repeated
OpenDP fits. Examine distributions of structural metrics rather than selecting
one favorable privacy-noise draw. The simulation-evaluation design and full
issue registry are in `design/TEST_SIM.md`.
