---
title: "Introduction to pmxSynthData"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Introduction to pmxSynthData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(pmxSynthData)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)

# Observed rows only, in a common shape, so five datasets with different
# schemas can be plotted and summarized together.
observations <- function(data, roles, label) {
  keep <- as.character(data[[roles$evid]]) %in% c("0", "0.0") &
    !is.na(data[[roles$dv]])
  if (!is.null(roles$mdv)) {
    keep <- keep & as.character(data[[roles$mdv]]) %in% c("0", "0.0")
  }
  data.frame(
    method = label,
    subject = as.character(data[[roles$id]][keep]),
    time = as.numeric(data[[roles$time]][keep]),
    dv = as.numeric(data[[roles$dv]][keep]),
    stringsAsFactors = FALSE
  )
}
```

## What this package does

`pmxSynthData` builds synthetic pharmacometric (PMX) datasets: dosing and
measurement event tables with the same schema, event grammar, and rough
behavior as a real study, so that data-assembly code, diagnostic plots, and
model-run plumbing can be developed outside the restricted environment that
holds the real data.

The hard part is not making numbers. It is deciding **how much of the real
data is allowed to survive into the synthetic data**, which is a privacy
question before it is a technical one. This package therefore offers four
generation modes that sit at different points on that scale. This vignette
introduces all four and applies each one to the same public dataset.

The four modes, from most faithful to most protective:

1. **AVATAR blending** — build each synthetic subject out of real subjects.
2. **Prior only** — read no data at all; simulate from a public model.
3. **Calibration** — simulate from a public model whose magnitude is corrected
   by a small, differentially private release.
4. **Empirical** — release a dense set of differentially private summaries and
   rebuild subjects from them.

Modes 2, 3, and 4 all use differential privacy (DP) accounting, and mode 2 is
its trivial limit: spending nothing.

## The example: theophylline

`theo_md` (from `nlmixr2data`) is a small public dataset — 12 subjects, seven
daily 320 mg oral doses, one concentration endpoint. Twelve subjects is a
deliberately hard case: it is typical of early-phase work and it is where the
differences between the four modes are most visible.

```{r data, eval=has_nlmixr2data}
data("theo_md", package = "nlmixr2data")
theo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", covariates = "WT"
)
str(theo_md)
```

## Mode 1: AVATAR blending

Start here, because it is the least work. `synthesize_pmx()` needs nothing but
the data and a declaration of what the columns mean. For each synthetic subject
it copies a real subject's event skeleton, then fills the covariates and
concentrations with a distance-weighted blend of that subject's nearest
compatible neighbors, plus noise.

```{r avatar, eval=has_nlmixr2data}
avatar <- suppressWarnings(synthesize_pmx(theo_md, theo_roles, seed = 101))
validate_pmx(avatar, theo_roles)$valid
```

Nothing was elicited, nothing was assumed, and the output keeps the source
schema and cohort size. On a 12-subject dataset `synthesize_pmx()` also emits
documented small-group fallback warnings (some event-pattern groups have only
one usable donor); they are suppressed above and explained in the "How
pmxSynthData Works" vignette.

What you cannot say about this output is that it is anonymous. It is assembled
from real trajectories, so it belongs inside the trusted environment that the
source data came from.

## Mode 2: prior only

The opposite extreme. Declare a public structural model and a public protocol,
and simulate. No confidential data is read, so there is nothing to protect and
no budget to spend: this is `epsilon = 0`, the strongest possible guarantee.

The typical parameter values must come from somewhere that is not the data —
allometric scaling from preclinical work, a published model for the compound
class, or the reasoning that set the starting dose.

```{r prior, eval=has_nlmixr2data}
theo_model <- pmx_structural_model(
  pk = "1cmt_oral",
  typical = c(cl = 6, v = 35, ka = 1.5),          # deliberately imperfect
  source = "illustrative allometric scaling; never fitted to theo_md"
)
theo_design <- pmx_trial_design(
  dose_levels = 320, cohort_sizes = 12,
  sampling = c(0, 0.25, 0.5, 1, 2, 4, 7, 9, 12, 24),
  n_doses = 7, dose_interval = 24,
  source = "illustrative protocol"
)
prior_only <- pmx_generate(theo_model, theo_design, n_subjects = 12, seed = 202)
head(prior_only, 3)
```

The generated table uses the package's own generated schema
(`pmx_generated_roles()`), including nominal time, time after dose, and
occasion columns. The clearance of 6 L/h assumed above is about twice the truth
for theophylline, and the output shows it: concentrations run low. That is the
honest cost of spending no budget — the data is exactly as good as the prior.

## Mode 3: calibration

The middle path, and the recommended one when a formal guarantee is needed and
the cohort is small. Keep the public model's *shape*, and spend a small privacy
budget correcting only its *magnitude*.

Each subject is reduced to a bounded multiplicative correction of the model's
own prediction, clipped to a public prior range, and released with calibrated
noise. Only two numbers leave the data: the correction and a noised subject
count.

```{r preflight, eval=has_nlmixr2data}
priors <- pmx_priors(pk = pmx_prior(c(1 / 4, 4), source = "scaling literature"))
pmx_preflight(priors, epsilon = 1, n_subjects = 12)
```

`pmx_preflight()` costs nothing and reads no data: it answers "is this release
worth its budget?" before any budget is spent.

```{r calibrate, eval=has_nlmixr2data}
calibrated <- fit_calibrated_pmx(
  data = theo_md, roles = theo_roles, model = theo_model,
  design = theo_design, priors = priors, epsilon = 1,
  backend = "public", public_source = TRUE   # theo_md is public; no DP claim
)
calibrated
calibrated_data <- pmx_generate(calibrated, seed = 303)
```

The correction pulled the assumed clearance toward the data. Generation is
post-processing on the released numbers, so drawing more datasets from one fit
spends no additional budget.

Two honest notes. First, `backend = "public"` is used here because `theo_md` is
already public: it makes the release **noiseless** and therefore makes no DP
claim, which the printed model states. A confidential fit uses the default
validated backend and fails closed if it is unavailable:

```{r opendp, eval=FALSE}
calibrated <- fit_calibrated_pmx(
  data = confidential, roles = theo_roles, model = theo_model,
  design = theo_design, priors = priors, epsilon = 1,
  backend = "opendp"
)
```

```{r backend, eval=requireNamespace("opendp", quietly = TRUE)}
dp_backend_status()
```

Second, at 12 subjects a genuine DP release of this correction is noisy enough
that it is often censored at the prior boundary — the package warns when that
happens, because the generated data then reflects the prior, not the study. The
"Exploring Epsilon" vignette covers when the release is worth making.

## Mode 4: empirical

The general-purpose private engine. Rather than asserting the curve shape, it
measures it: it releases noised summaries for the subject count, event and
regimen structure, observation timing, endpoint trajectories, baseline
covariates, and censoring, then rebuilds subjects from those summaries. This
buys realism that the public model does not contain, and pays for it by
splitting one epsilon across many released quantities.

It also needs the most declaration: every clipping range, contribution limit,
and budget share is an explicit public input.

```{r empirical, eval=has_nlmixr2data}
empirical <- fit_private_pmx(
  data = theo_md, roles = theo_roles,
  endpoints = list(cp = pmx_endpoint(
    alignment = "dose_relative", transform = "log", shape = "occasion", cmt = 2
  )),
  epsilon = 5, delta = 0,
  bounds = pmx_bounds(
    time = c(0, 170), endpoints = list(cp = c(0, 30)), amt = c(0, 500),
    covariates = list(WT = c(40, 130))
  ),
  public_design = pmx_public_design(
    pmx_schema(theo_md), dose_evid = 101, dose_cmt = 1
  ),
  contribution_limits = pmx_contribution_limits(40, 8, 8, 30, 11),
  budget_allocation = pmx_budget_allocation(
    subject_count = 0.10, event = 0.15, timing = 0.15,
    covariates = 0.10, endpoints = 0.50, censoring = 0
  ),
  backend = "public", public_source = TRUE   # theo_md is public; no DP claim
)
empirical_data <- generate_pmx(empirical, seed = 404)
privacy_report(empirical)
```

Note the requested `epsilon = 5`, five times the calibrated fit's budget, for a
worse result at this cohort size. That is not a bug: the same budget is being
split six ways over dozens of released coordinates. This engine earns its keep
on large pooled corpora, not on twelve subjects.

## The four side by side

```{r compare, eval=has_nlmixr2data}
generated_roles <- pmx_generated_roles()
all_observations <- rbind(
  observations(theo_md, theo_roles, "Source"),
  observations(avatar, theo_roles, "1. AVATAR"),
  observations(prior_only, generated_roles, "2. Prior only"),
  observations(calibrated_data, generated_roles, "3. Calibration"),
  # The empirical engine restores the source schema, so it uses source roles.
  observations(empirical_data, theo_roles, "4. Empirical")
)
summaries <- do.call(rbind, lapply(
  split(all_observations$dv, all_observations$method),
  function(dv) {
    data.frame(
      n_observations = length(dv),
      median = stats::median(dv),
      p10 = stats::quantile(dv, 0.10, names = FALSE),
      p90 = stats::quantile(dv, 0.90, names = FALSE)
    )
  }
))
knitr::kable(
  summaries, digits = 2,
  caption = "Observed concentrations by generation mode"
)
```

```{r compare-plot, eval=has_nlmixr2data && has_ggplot2, fig.height=6}
ggplot2::ggplot(
  all_observations,
  ggplot2::aes(time, dv, group = subject)
) +
  ggplot2::geom_line(alpha = 0.4, colour = "#1B6CA8") +
  ggplot2::geom_point(alpha = 0.5, size = 0.7, colour = "#1B6CA8") +
  ggplot2::facet_wrap(~ method, ncol = 2) +
  ggplot2::labs(
    x = "Study time (hours)", y = "Concentration",
    title = "One dataset, four generation modes"
  ) +
  ggplot2::theme_minimal()
```

AVATAR tracks the source most closely, because it is made of it. The
prior-only data has the right structure and the wrong level. Calibration moves
the level toward the truth for a small budget. The empirical release recovers
more of the real timing and spread, but at this cohort size the noise is
visible.

## Choosing a mode

```{r modes-table, echo=FALSE}
modes <- data.frame(
  Mode = c("1. AVATAR blending", "2. Prior only", "3. Calibration",
           "4. Empirical"),
  Function = c("synthesize_pmx()", "pmx_generate(model, design)",
               "fit_calibrated_pmx() + pmx_generate()",
               "fit_private_pmx() + generate_pmx()"),
  `Output built from` = c(
    "Real subject templates and blended real trajectories",
    "A public model and protocol only",
    "A public model, magnitude corrected by 2 private releases",
    "Dozens of noised population summaries"
  ),
  Guarantee = c(
    "None; governance only",
    "epsilon = 0 (no data read)",
    "(epsilon, delta) DP",
    "(epsilon, delta) DP"
  ),
  `Cohort size` = c("Any, from ~12", "Any (data-independent)",
                    "~20 and up", "A few hundred and up"),
  `Elicitation needed` = c(
    "None", "Structural model + protocol",
    "Model, protocol, prior ranges",
    "Endpoints, bounds, limits, budget split"
  ),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(modes)
```

Where each mode belongs:

| Environment | Appropriate modes | Why |
|---|---|---|
| Inside the validated environment holding the source data; you are the only consumer | **AVATAR**, or any other | Access control and governance already bound the risk. A formal guarantee defends against an adversary who cannot reach the output, so it buys nothing and costs utility. |
| Shared with a partner, vendor, or contract research organization (CRO) | **Calibration** or **Empirical**, with an approved epsilon | The output leaves your controls. A contract is not a mathematical bound; DP is what survives a determined recipient. |
| Published, posted to a repository, or shipped inside a package or teaching material | **Prior only**, or **Calibration** with a small approved epsilon | Anyone may inspect it, forever, alongside side information you cannot anticipate. Prior-only data reads no patient record at all and is the safest thing to publish. |
| Software testing where only schema and event grammar matter | **Prior only** | Fidelity is irrelevant; a data-independent generator removes the question entirely. |

Two rules of thumb behind the table:

- **The trust boundary decides whether you need DP.** Ask whether the generated
  data can reach anyone the source data could not. If not, AVATAR is more
  useful and its lack of a formal guarantee costs nothing. If so, only an
  accounted release holds up.
- **The cohort size decides which DP mode is usable.** Epsilon buys accuracy in
  proportion to the number of subjects and in inverse proportion to how many
  quantities you release. At 12 subjects, releasing two numbers can work and
  releasing fifty cannot.

Epsilon and delta are governance decisions, not defaults. For anything public
facing they should be set and justified by whoever owns the data, and recorded:
every fit carries a release ledger, and `privacy_report()` prints the realized
accounting.

## Where to go next

- **"Using pmxSynthData"** — the practical workflow across five public
  datasets, with structural checks.
- **"How pmxSynthData Works"** — the AVATAR algorithm in full, and the
  model-based alternatives.
- **"Privacy in pmxSynthData"** — what differential privacy guarantees, what it
  does not, and the trust-boundary decision rule.
- **"Exploring Epsilon in pmxSynthData"** — how epsilon trades against utility,
  and the measured frontier by cohort size.
- `design/METHOD_DISCUSSION.md` — why AVATAR is the default and DP the
  boundary-crossing alternative.
